Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States.

### 2017-02-03 Manuscript/Manuscript.Rmd

**Background notes from meeting with SH on Dec 20**
-CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.
-Richard cooper also written about BW diff in CVD with epi perspective
-for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides
-massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide
-fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation. Does this help us understand why the gap in Wisconsin was increasing?
-is there a lot of heterogeneity or is there not? - same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.
-heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.
-this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).
-then have a view of how the different COD have contributed to the decline in the gap over time.

@@

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancies between blacks and white was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.

To decompose the black-white difference in life expectancy, we performed a decomposition analysis using the Arriaga method [ref]. First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.

>>>>

**Background notes from meeting with SH on Dec 20**
- CVD declining starting 1968 - see paper Ruel Stallones - cited in SH JAMA paper - history of decline in CVD.

- Richard cooper also written about BW diff in CVD with epi perspective

- for men there is stagnation from 1969 to late 1980s and then increase in gap with HIV and homicides

- massive crime wave in 1980s, black LE went down from 1984 to 1987 - period when both HIV and homicide went up -- the whole story during that time was HIV and homicide

- fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation. Does this help us understand why the gap in Wisconsin was increasing?

- is there a lot of heterogeneity or is there not? - same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.

- heterogeneity between states in the contribution of the causes to the decline in the gap - maybe need to break up into at least two time periods with split at 1990.

- this is where the new figures come in which on the y-axis have the contribution to LE gap (in years or %) vs year on the x axis, and geom_line for each COD (do one COD at a time), and individual line for each state (may facet by census region).

- then have a view of how the different COD have contributed to the decline in the gap over time.

@@

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancies between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.

To decompose the year- and state- specific difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.

In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013). While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (Appendix Figures 1 and 2).
